---
title: "Spotify-Top-Songs-EDA"
author: "Riley Modlin"
date: "2025-01-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(dplyr)

DATA = read_csv("C:/Users/rimod/OneDrive/Desktop/spotify-top-songs-data-analysis/playlist_2010to2023.csv", col_names=T)

```

```{r}
DATA
```


Q1: Is there a relationship between artist_popularity and danceability?

```{r}
ggplot(DATA, aes(artist_popularity, danceability)) + geom_boxplot()

ggplot(DATA, aes(artist_popularity, danceability, color = year)) + geom_point()
```

Q2: Which years had the shortest/longest songs on average?

```{r}
duration_averages = DATA %>% 
                      group_by(year) %>%
                      summarize(avg_ms = mean(duration_ms))

duration_averages$avg_second <- duration_averages$avg_ms/1000

duration_averages$avg_min <- duration_averages$avg_second/60

duration_averages

ggplot(duration_averages, aes(year, avg_min)) + geom_point() +
geom_line(color = 'blue')

ggplot(duration_averages, aes(year, avg_min)) + geom_col()
```

Q3: How does song duration vary from year to year?

```{r}
DATA_grouped <- DATA %>%
  mutate(year_group = case_when(
    year < 2005 ~ "2000-2004",
    year >= 2005 & year < 2010 ~ "2005-2009",
    year >= 2010 & year < 2015 ~ "2010-2014",
    year >= 2015 & year < 2020 ~ "2015-2019",
    year >= 2020 ~ "2020-2023"
  ))

ggplot(DATA_grouped, aes(x = year, y = duration_ms, color = year_group)) +
  geom_point(size = 2) +  # Adjust point size
  scale_color_manual(values = c("2000-2004" = "#6A69BF", 
                                "2005-2009" = "#84EEF0", 
                                "2010-2014" = "#F071E8", 
                                "2015-2019" = "#61D6A1", 
                                "2020-2023" = "#FFB957"))


```

Q4: Is there a relationship between artist_popularity and track_popularity?

```{r}
ggplot(DATA, aes(artist_popularity, track_popularity)) + geom_boxplot()

ggplot(DATA_grouped, aes(x = artist_popularity, y = track_popularity, color = year_group)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("2000-2004" = "#6A69BF", 
                                "2005-2009" = "#84EEF0", 
                                "2010-2014" = "#F071E8", 
                                "2015-2019" = "#61D6A1", 
                                "2020-2023" = "#FFB957"))
```

Q5: Is there a relationship between speechiness and acousticness?

```{r}
ggplot(DATA, aes(speechiness, acousticness)) + geom_boxplot()

ggplot(DATA_grouped, aes(x = speechiness, y = acousticness, color = year_group)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("2000-2004" = "#6A69BF", 
                                "2005-2009" = "#84EEF0", 
                                "2010-2014" = "#F071E8", 
                                "2015-2019" = "#61D6A1", 
                                "2020-2023" = "#FFB957"))
```

Q6: What genres were common among the most popular songs?

```{r}
DATA_most_popular <- DATA_grouped[DATA_grouped$track_popularity >= 90, , drop = FALSE]

DATA_most_popular
```

