---
title: "Spotify-Top-Songs-EDA"
author: "Riley Modlin"
date: "2025-01-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(dplyr)

DATA = read_csv("C:/Users/rimod/OneDrive/Desktop/spotify-top-songs-data-analysis/playlist_2010to2023.csv", col_names=T)

```

```{r}
DATA

DATA_grouped <- DATA %>%
  mutate(year_group = case_when(
    year < 2005 ~ "2000-2004",
    year >= 2005 & year < 2010 ~ "2005-2009",
    year >= 2010 & year < 2015 ~ "2010-2014",
    year >= 2015 & year < 2020 ~ "2015-2019",
    year >= 2020 ~ "2020-2023"
  ))
```

Q1: Is there a relationship between artist_popularity and danceability?

```{r}
ggplot(DATA_grouped, aes(as.character(year), artist_popularity)) + geom_boxplot()

ggplot(DATA_grouped, aes(as.character(year), danceability)) + geom_boxplot()

DATA_grouped %>% 
  group_by(year) %>%
  summarize(avg_artist_popularity = mean(artist_popularity), avg_danceability = mean(danceability))

ggplot(DATA_grouped, aes(artist_popularity, danceability, color = year_group)) + geom_point() +
  scale_color_manual(values = c("2000-2004" = "#242442", 
                                "2005-2009" = "#414176", 
                                "2010-2014" = "#57579E", 
                                "2015-2019" = "#7B7BB7", 
                                "2020-2023" = "#A3A3CC"))

ggplot(DATA_grouped, aes(artist_popularity, danceability)) + geom_smooth(color = "#7B7BB7", se = F)
```

Q2: Which years had the shortest/longest songs on average?

```{r}
duration_averages = DATA %>% 
                      group_by(year) %>%
                      summarize(avg_ms = mean(duration_ms))

duration_averages$avg_second <- duration_averages$avg_ms/1000

duration_averages$avg_min <- duration_averages$avg_second/60

duration_averages

ggplot(duration_averages, aes(year, avg_min)) + geom_point() +
geom_line(color = '#7B7BB7')

ggplot(duration_averages, aes(year, avg_min)) + geom_col(fill = "#7B7BB7")
```

Q3: How does song duration vary from year to year?

```{r}

bins <- sqrt(2400)

bw <- 2 * IQR(DATA_grouped$duration_ms) / length(DATA_grouped$duration_ms)^(1/3)

ggplot(DATA_grouped, aes(x = duration_ms)) + geom_histogram(fill= "#7B7BB7", bins = bins, binwidth = bw) + facet_wrap(~year)

```

Q4: Is there a relationship between artist_popularity and track_popularity?

```{r}
ggplot(DATA, aes(as.character(year), artist_popularity)) + geom_boxplot()

ggplot(DATA_grouped, aes(as.character(year), track_popularity)) + geom_boxplot()

DATA_grouped %>% 
  group_by(year) %>%
  summarize(avg_artist_popularity = mean(artist_popularity), avg_track_popularity = mean(track_popularity))

ggplot(DATA_grouped, aes(x = artist_popularity, y = track_popularity, color = year_group)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("2000-2004" = "#242442", 
                                "2005-2009" = "#414176", 
                                "2010-2014" = "#57579E", 
                                "2015-2019" = "#7B7BB7", 
                                "2020-2023" = "#A3A3CC"))

ggplot(DATA_grouped, aes(artist_popularity, danceability)) + geom_smooth(color = "#7B7BB7", se = F)
  
```

Q5: Is there a relationship between speechiness and acousticness?

```{r}
ggplot(DATA, aes(as.character(year), speechiness)) + geom_boxplot()

ggplot(DATA_grouped, aes(as.character(year), acousticness)) + geom_boxplot()

DATA_grouped %>% 
  group_by(year) %>%
  summarize(avg_speechiness = mean(speechiness), avg_acousticness = mean(acousticness))

ggplot(DATA_grouped, aes(x = speechiness, y = acousticness, color = year_group)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("2000-2004" = "#242442", 
                                "2005-2009" = "#414176", 
                                "2010-2014" = "#57579E", 
                                "2015-2019" = "#7B7BB7", 
                                "2020-2023" = "#A3A3CC"))

ggplot(DATA_grouped, aes(x = speechiness, y = acousticness)) + geom_smooth(color = "#7B7BB7", se = F)
```

Q6: What genres were common among the most popular songs?

```{r}
DATA_most_popular <- DATA_grouped[DATA_grouped$track_popularity >= 90, , drop = FALSE]

unique_songs <- DATA_most_popular %>% distinct(track_name, .keep_all = TRUE)

unique_songs

genres <- c('permanent wave','pop','dance pop','pop','post-teen pop',	
'dance pop','pop','chill pop','canadian contemporary r&b','canadian pop','pop',
'pop','pop','canadian contemporary r&b','canadian pop','pop','big room','dance pop','edm','pop', 'pop dance','art pop','pop')

genre_data <- data.frame(Genre = genres)

ggplot(genre_data, aes(x = Genre)) +
  geom_bar(fill = "#7B7BB7") + 
  labs(title = "Frequency of Song Genres", x = "Genre", y = "Count") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Q7: What genres were most popular during each year?

```{r}
genre_counts <- DATA %>%
  group_by(year, artist_genres) %>%
  summarise(Count = n(), .groups = 'drop')

most_popular_genres <- genre_counts %>%
  group_by(year) %>%
  filter(Count == max(Count)) %>%
  ungroup()

most_popular_genres
```

Q8: Is there a relationship between track_popularity and variables such as (danceability, energy, key, loudness, mode, speechiness, acousticness, instrumentalness, liveness, valence, tempo, duration_ms, time_signature) ?

```{r}
#track_popularity & danceability

ggplot(DATA_grouped, aes(x = track_popularity, y = danceability, color = year_group)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("2000-2004" = "#242442", 
                                "2005-2009" = "#414176", 
                                "2010-2014" = "#57579E", 
                                "2015-2019" = "#7B7BB7", 
                                "2020-2023" = "#A3A3CC"))

ggplot(DATA_grouped, aes(x = track_popularity, y = danceability)) + geom_smooth(color = "#7B7BB7", se = F)

#track_popularity & energy

ggplot(DATA_grouped, aes(x = track_popularity, y = energy, color = year_group)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("2000-2004" = "#242442", 
                                "2005-2009" = "#414176", 
                                "2010-2014" = "#57579E", 
                                "2015-2019" = "#7B7BB7", 
                                "2020-2023" = "#A3A3CC"))

ggplot(DATA_grouped, aes(x = track_popularity, y = energy)) + geom_smooth(color = "#7B7BB7", se = F)

#track_popularity & key

ggplot(DATA_grouped, aes(x = track_popularity, y = key, color = year_group)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("2000-2004" = "#242442", 
                                "2005-2009" = "#414176", 
                                "2010-2014" = "#57579E", 
                                "2015-2019" = "#7B7BB7", 
                                "2020-2023" = "#A3A3CC"))

ggplot(DATA_grouped, aes(x = track_popularity, y = key)) + geom_smooth(color = "#7B7BB7", se = F)

#track_popularity & loudness

ggplot(DATA_grouped, aes(x = track_popularity, y = loudness, color = year_group)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("2000-2004" = "#242442", 
                                "2005-2009" = "#414176", 
                                "2010-2014" = "#57579E", 
                                "2015-2019" = "#7B7BB7", 
                                "2020-2023" = "#A3A3CC"))

ggplot(DATA_grouped, aes(x = track_popularity, y = loudness)) + geom_smooth(color = "#7B7BB7", se = F)

#track_popularity & mode

ggplot(DATA_grouped, aes(x = track_popularity, y = mode, color = year_group)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("2000-2004" = "#242442", 
                                "2005-2009" = "#414176", 
                                "2010-2014" = "#57579E", 
                                "2015-2019" = "#7B7BB7", 
                                "2020-2023" = "#A3A3CC"))

ggplot(DATA_grouped, aes(x = track_popularity, y = mode)) + geom_smooth(color = "#7B7BB7", se = F)

#track_popularity & speechiness

ggplot(DATA_grouped, aes(x = track_popularity, y = speechiness, color = year_group)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("2000-2004" = "#242442", 
                                "2005-2009" = "#414176", 
                                "2010-2014" = "#57579E", 
                                "2015-2019" = "#7B7BB7", 
                                "2020-2023" = "#A3A3CC"))

ggplot(DATA_grouped, aes(x = track_popularity, y = speechiness)) + geom_smooth(color = "#7B7BB7", se = F)

#track_popularity & acousticness

ggplot(DATA_grouped, aes(x = track_popularity, y = acousticness, color = year_group)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("2000-2004" = "#242442", 
                                "2005-2009" = "#414176", 
                                "2010-2014" = "#57579E", 
                                "2015-2019" = "#7B7BB7", 
                                "2020-2023" = "#A3A3CC"))

ggplot(DATA_grouped, aes(x = track_popularity, y = acousticness)) + geom_smooth(color = "#7B7BB7", se = F)

#track_popularity & instrumentalness

ggplot(DATA_grouped, aes(x = track_popularity, y = instrumentalness, color = year_group)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("2000-2004" = "#242442", 
                                "2005-2009" = "#414176", 
                                "2010-2014" = "#57579E", 
                                "2015-2019" = "#7B7BB7", 
                                "2020-2023" = "#A3A3CC"))

ggplot(DATA_grouped, aes(x = track_popularity, y = instrumentalness)) + geom_smooth(color = "#7B7BB7", se = F)

#track_popularity & liveness

ggplot(DATA_grouped, aes(x = track_popularity, y = liveness, color = year_group)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("2000-2004" = "#242442", 
                                "2005-2009" = "#414176", 
                                "2010-2014" = "#57579E", 
                                "2015-2019" = "#7B7BB7", 
                                "2020-2023" = "#A3A3CC"))

ggplot(DATA_grouped, aes(x = track_popularity, y = liveness)) + geom_smooth(color = "#7B7BB7", se = F)

#track_popularity & valence

ggplot(DATA_grouped, aes(x = track_popularity, y = valence, color = year_group)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("2000-2004" = "#242442", 
                                "2005-2009" = "#414176", 
                                "2010-2014" = "#57579E", 
                                "2015-2019" = "#7B7BB7", 
                                "2020-2023" = "#A3A3CC"))

ggplot(DATA_grouped, aes(x = track_popularity, y = valence)) + geom_smooth(color = "#7B7BB7", se = F)

#track_popularity & tempo

ggplot(DATA_grouped, aes(x = track_popularity, y = tempo, color = year_group)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("2000-2004" = "#242442", 
                                "2005-2009" = "#414176", 
                                "2010-2014" = "#57579E", 
                                "2015-2019" = "#7B7BB7", 
                                "2020-2023" = "#A3A3CC"))

ggplot(DATA_grouped, aes(x = track_popularity, y = tempo)) + geom_smooth(color = "#7B7BB7", se = F)

#track_popularity & duration_ms

ggplot(DATA_grouped, aes(x = track_popularity, y = duration_ms, color = year_group)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("2000-2004" = "#242442", 
                                "2005-2009" = "#414176", 
                                "2010-2014" = "#57579E", 
                                "2015-2019" = "#7B7BB7", 
                                "2020-2023" = "#A3A3CC"))

ggplot(DATA_grouped, aes(x = track_popularity, y = duration_ms)) + geom_smooth(color = "#7B7BB7", se = F)

#track_popularity & time_signature

ggplot(DATA_grouped, aes(x = track_popularity, y = time_signature, color = year_group)) +
  geom_point(size = 1) +
  scale_color_manual(values = c("2000-2004" = "#242442", 
                                "2005-2009" = "#414176", 
                                "2010-2014" = "#57579E", 
                                "2015-2019" = "#7B7BB7", 
                                "2020-2023" = "#A3A3CC"))

ggplot(DATA_grouped, aes(x = track_popularity, y = time_signature)) + geom_smooth(color = "#7B7BB7", se = F)

#averages

DATA_grouped %>% 
  group_by(year) %>%
  summarize(avg_track_popularity = mean(track_popularity), avg_danceability = mean(danceability), avg_energy = mean(energy), avg_loudness = mean(loudness), avg_speechiness = mean(speechiness), avg_acousticness = mean(acousticness), avg_instrumentalness = mean(instrumentalness), avg_liveness = mean(liveness), avg_valence = mean(valence), avg_tempo = mean(tempo), avg_duration_ms = mean(duration_ms))

```
